%% thesis.tex 2014/04/11
%
% Based on sample files of unknown authorship.
%
% The Current Maintainer of this work is Paul Vojta.

\documentclass[oldfontcommands]{ucbthesis}
\usepackage{biblatex}
\usepackage{amsmath, amssymb}
\usepackage{times}
\usepackage{graphicx}
\graphicspath{ {images/} }
\usepackage{color}
\usepackage{multirow}
\usepackage{rotating}
\usepackage{bbm}
\usepackage{latexsym}

% To compile this file, run "latex thesis", then "biber thesis"
% (or "bibtex thesis", if the output from latex asks for that instead),
% and then "latex thesis" (without the quotes in each case).

% Double spacing, if you want it.  Do not use for the final copy.
% \def\dsp{\def\baselinestretch{2.0}\large\normalsize}
% \dsp

% If the Grad. Division insists that the first paragraph of a section
% be indented (like the others), then include this line:
% \usepackage{indentfirst}

\newtheorem{theorem}{Jibberish}

\bibliography{references}

\hyphenation{mar-gin-al-ia}
\hyphenation{bra-va-do}

\begin{document}

% Declarations for Front Matter

\title{Neural Representation Learning with Denoising Autoencoder Framework}
\author{Chayut Thanapirom}
\degreesemester{Spring}
\degreeyear{2016}
\degree{Doctor of Philosophy}
\chair{Professor Michael R. DeWeese}
\othermembers{Professor Bruno A. Olshausen \\
  Professor Hernan G. Garcia}
\numberofmembers{3}
% Previous degrees are no longer to be listed on the title page.
% \prevdegrees{B.A. (University of Northern South Dakota at Hoople) 1978 \\
%   M.S. (Ed's School of Quantum Mechanics and Muffler Repair) 1989}
\field{Physics}
% Designated Emphasis -- this is optional, and rare
% \emphasis{Colloidal Telemetry}
% This is optional, and rare
% \jointinstitution{University of Western Maryland}
% This is optional
\campus{Berkeley}

% For a masters thesis, replace the above \documentclass line with
% \documentclass[masters]{ucbthesis}
% This affects the title and approval pages, which by default calls this
% document a "dissertation", not a "thesis".

\maketitle
% Delete (or comment out) the \approvalpage line for the final version.
%\approvalpage
\copyrightpage

\include{abstract}

\begin{frontmatter}

\begin{dedication}
\null\vfil
\begin{center}
In memory of my mother and my grandfather. \\\vspace{12pt}
\end{center}
\vfil\null
\end{dedication}

% You can delete the \clearpage lines if you don't want these to start on
% separate pages.

\tableofcontents
\clearpage
\listoffigures
\clearpage
\listoftables

\begin{acknowledgements}
First and foremost, I would like to thank my advisor Mike DeWeese, for working with me, and providing me with support and encouragement. He always believes in me even when I do not really believe in myself.

Secondly, thank you to my co-adviser, Bruno Olshausen. Bruno spends a lot of his time formulating ideas with me and helping me out when I face problems, even when he is busy. Also, thank you for being my academic role-model.

I also owe a huge debt to Hernan Garcia for being on my thesis committee. Thank you so much for donating your time to help me graduate.

Big thanks to everyone in the Redwood Center for listening to my presentations and giving me constructive feedbacks, for helping me practice my qual exam, for all the discussion we have, for all the support, and for making the Redwood Center, in my opinion, the best place to work in Theoretical Neuroscience.

Thank you to Chanoknan Vorasucha for helping me write some of the code for running the experiments. 

Thank you to my family and my friends for all the mental support through the difficult times. I do not believe I can get through graduate school without them.

Lastly, I gratefully acknowledge the support of NVIDIA Corporation with the donation of the Tesla K40 GPU used for this research.

\end{acknowledgements}

\end{frontmatter}

\pagestyle{headings}

% (Optional) \part{First Part}

\include{chap1}
\include{chap2}
\include{chap3}
\include{chap4}
\include{appendix}

\printbibliography

\end{document}
